Toward nonlinear multigrid \\ for nonlinear and nonlocal variational inequalities

### 2023-02-08 talk/references.tex
more talk work ... more work todo
fast and furious

\newcommand{\sdoi}[1]{\,{\tiny \href{https://doi.org/#1}{doi:#1}}}
\begin{itemize}
\item[] A.~Brandt \& C.~Cryer (1983). \emph{Multigrid algorithms for the solution of linear complementarity problems \dots}, SIAM J.~Sci.~Stat.~Comput.~4 (4), 655--684 \sdoi{10.1137/0904046}
%FTITLE Multigrid algorithms for the solution of linear complementarity problems arising from free boundary problems
\item[] P.~Brune, M.~Knepley, B.~Smith, \& X.~Tu (2015). \emph{Composing scalable nonlinear algebraic solvers}, SIAM Review~57 (4), 535--565 \sdoi{10.1137/130936725}
\item[] E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032 \sdoi{10.1137/20M135217X}
%\item[] E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \sdoi{10.1017/jog.2022.113}
\item[] G.~de Diego, P.~Farrell, \& I.~Hewitt (2022). \emph{Numerical approximation of viscous contact problems applied to glacial sliding}, J.~Fluid Mech.~938 (A21). \sdoi{10.1017/jfm.2022.178}
\item[] P.~Farrell, M.~Croci, \& T.~Surowiec (2020). \emph{Deflation for semismooth equations}, Optimization Methods \& Software 35 (6), 1248--1271 \sdoi{10.1080/10556788.2019.1613655}
\item[] C.~Gr{\"a}ser \& R.~Kornhuber (2009). \emph{Multigrid methods for obstacle problems}, J.~Comput.~Math., 1--44
\item[] T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations \dots ice sheet dynamics}, SIAM J.~Sci.~Comput., 37 (6), B804--B833 \sdoi{10.1137/140974407}
%FTITLE Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics

\item[] G. Jouvet \& E. Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems \dots}, SIAM J.~Appl.~Math.~72 (4), 1292--1314 \sdoi{10.1137/110856654}
%FTITLE Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation
\item[] A.~Reusken (1987). \emph{Convergence of the multigrid full approximation scheme \dots}, Numer.~Math.~52, 251--277 \sdoi{10.1007/BF01398879}
%FTITLE Convergence of the multigrid full approximation scheme for a class of elliptic mildly nonlinear boundary value problems
\item[] J.~Seguinot \& 5 others (2018).  \emph{Modelling last glacial cycle ice dynamics in the Alps}, The Cryosphere, 12 (10), 3265--3285 \sdoi{10.5194/tc-12-3265-2018}
%FAUTHORS Seguinot, J., Ivy-Ochs, S., Jouvet, G., Huss, M., Funk, M., & Preusser, F.
\end{itemize}

>>>>

\newcommand{\sdoi}[1]{\,{\tiny \href{https://doi.org/#1}{doi:#1}}}
\begin{itemize}
\item A.~Brandt \& C.~Cryer (1983). \emph{Multigrid algorithms for the solution of linear complementarity problems \dots}, SIAM J.~Sci.~Stat.~Comput.~4 (4), 655--684 \sdoi{10.1137/0904046}
%FTITLE Multigrid algorithms for the solution of linear complementarity problems arising from free boundary problems
\item P.~Brune, M.~Knepley, B.~Smith, \& X.~Tu (2015). \emph{Composing scalable nonlinear algebraic solvers}, SIAM Review~57 (4), 535--565 \sdoi{10.1137/130936725}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032 \sdoi{10.1137/20M135217X}
%\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \sdoi{10.1017/jog.2022.113}
\item G.~de Diego, P.~Farrell, \& I.~Hewitt (2022). \emph{Numerical approximation of viscous contact problems applied to glacial sliding}, J.~Fluid Mech.~938 (A21). \sdoi{10.1017/jfm.2022.178}
\item P.~Farrell, M.~Croci, \& T.~Surowiec (2020). \emph{Deflation for semismooth equations}, Optimization Methods \& Software 35 (6), 1248--1271 \sdoi{10.1080/10556788.2019.1613655}
\item C.~Gr{\"a}ser \& R.~Kornhuber (2009). \emph{Multigrid methods for obstacle problems}, J.~Comput.~Math., 1--44
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations \dots ice sheet dynamics}, SIAM J.~Sci.~Comput., 37 (6), B804--B833 \sdoi{10.1137/140974407}
%FTITLE Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics
\item G.~Jouvet \& E.~Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems \dots}, SIAM J.~Appl.~Math.~72 (4), 1292--1314 \sdoi{10.1137/110856654}
%FTITLE Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation
\item G.~Jouvet \& J.~Rappaz (2011). \emph{Analysis and finite element approximation of a nonlinear stationary {S}tokes problem \dots}, Adv.~Numer.~Analysis 2011 (164581) \sdoi{10.1155/2011/164581}
%FTITLE Analysis and finite element approximation of a nonlinear stationary {S}tokes problem arising in glaciology
\item R.~Kornhuber (1994). \emph{Monotone multigrid methods for elliptic variational inequalities I}, Numer.~Math.~69, 167--184 \sdoi{10.1007/BF03325426}
\item A.~Reusken (1987). \emph{Convergence of the multigrid full approximation scheme \dots}, Numer.~Math.~52, 251--277 \sdoi{10.1007/BF01398879}
%FTITLE Convergence of the multigrid full approximation scheme for a class of elliptic mildly nonlinear boundary value problems
\item J.~Seguinot \& 5 others (2018).  \emph{Modelling last glacial cycle ice dynamics in the Alps}, The Cryosphere, 12 (10), 3265--3285 \sdoi{10.5194/tc-12-3265-2018}
%FAUTHORS Seguinot, J., Ivy-Ochs, S., Jouvet, G., Huss, M., Funk, M., & Preusser, F.
\item X.~Tai (2003). \emph{Rate of convergence for some constraint decomposition methods for nonlinear variational inequalities}, Numer.~Math.~93 (4), 755--786 \sdoi{10.1007/s002110200404}
\end{itemize}

### 2023-02-08 talk/slides.tex
fast and furious

%\usepackage[svgnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
    pdftitle = {Toward nonlinear multigrid for nonlinear and nonlocal variational inequalities},
    pdfauthor = {Ed Bueler},
    pdfsubject = {},
    pdfkeywords = {}
}

@@

\title[Multigrid for nonlinear VI]{Toward nonlinear multigrid \\ for nonlinear and non-local variational inequalities}

@@

\author{Ed Bueler}

\institute[UAF]{University of Alaska Fairbanks}

@@

\titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{images/uafbw.png}}

@@

\section{the nonlocal VI for a fluid layer in a climate}

@@

    \begin{itemize}
    \item[$\circ$] $\mathcal{V}$ to be determined by viscous fluid model\footnote{in shallow ice approximation, $(s-b)^{8/3} \in W^{1,4}(\Omega)$ (Jouvet \& Bueler, 2012)}
    \end{itemize}
\item VI problem for surface elevation $s\in\mathcal{K}$:
	$$\ip{\Phi(s)}{r-s} \ge \ip{a}{r-s} \quad \text{ for all } r \in \mathcal{K}$$
where
    $$\Phi(s)=- \bu|_s \cdot \bn_s,$$
with extension by 0 to all of $\Omega$, and $\bu$ is the velocity solution on
    $$\Lambda(s) = \{(x,y,z) : b(x,y) < z < s(x,y)\}$$

\end{itemize}
\end{frame}

@@

\section{multigrid approaches for VIs?}

\begin{frame}[fragile]
\frametitle{Newton-multigrid for VIs}

\begin{itemize}
\item multigrid for the classical obstacle problem?
    \begin{itemize}
    \item[$\circ$] a nonlinear problem, though $-\nabla^2$ is linear
    \end{itemize}
\item Newton-multigrid is straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg \
    -pc_type mg -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] CG-GMG V-cycles linear solver applies to inactive variables
    \end{itemize}
\item the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] thus growing Newton (SNES) iterations!
    \end{itemize}
\end{itemize}

\hfill \mbox{\includegraphics[height=0.2\textheight]{images/obstacle65.pdf} \hspace{15mm}
\includegraphics[height=0.25\textheight]{images/vi-newton-gmg-bad.png}}
\end{frame}

\begin{frame}[fragile]
\frametitle{Newton-multigrid for VIs: nested iteration}

@@

\vspace{2mm}
    \begin{itemize}
    \item[$\circ$] constant Newton (SNES) iterations
    \item[$\circ$] grid-independent flops/time restored
    \item[$\circ$] optimal $O(N^1)$ flops
    \item[$\circ$] FIXME example in my book
    \end{itemize}

@@

\bigskip
\hspace{10mm} \includegraphics[height=0.25\textheight]{images/vi-newton-gmg-good.png}

\vspace{-22mm}
\hfill \includegraphics[height=0.45\textheight]{images/obstacle-flops-per-n.png}
\end{frame}

\begin{frame}{Newton-multigrid for VIs: semi-smooth}

xxx

semi-smooth Newton with penalty

scaling also restores optimal

performance (Farrell et al.~2020)

FIXME Ulrich (20xx)

\end{frame}

\section{multigrid approaches for \emph{nonlocal} VIs?}

@@

{\scriptsize
%{\notsotiny
\input{references.tex}
}
\end{frame}

@@

{\scriptsize
%{\notsotiny
\input{background.tex}
}
\end{frame}

>>>>

%\usepackage[svgnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
    pdftitle = {Toward nonlinear multigrid for nonlinear and nonlocal variational inequalities},
    pdfauthor = {Ed Bueler and Patrick Farrell},
    pdfsubject = {},
    pdfkeywords = {}
}

@@

\title[Multigrid for nonlinear and nonlocal VIs]{Toward nonlinear multigrid \\ for nonlinear and nonlocal variational inequalities}

@@

\author[Bueler and Farrell]{Ed Bueler \inst{1} \and Patrick Farrell \inst{2}}
\institute[]{\inst{1} University of Alaska Fairbanks \and %
             \inst{2} Mathematical Institute, Oxford University}

@@

\titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{images/uafbw.png} \quad \includegraphics[width=0.10\textwidth]{images/oxford.png}}

@@

\section{the nonlinear and nonlocal VI for a fluid layer in a climate}

@@

    \begin{itemize}
    \item[$\circ$] $\mathcal{V}$ to be determined by viscous fluid model\footnote{in shallow ice approximation, $(s-b)^{8/3} \in W^{1,4}(\Omega)$ (Jouvet \& Bueler, 2012)}
    \end{itemize}
\item VI problem for surface elevation $s\in\mathcal{K}$:
	$$\ip{\Phi(s)}{r-s} \ge \ip{a}{r-s} \quad \text{ for all } r \in \mathcal{K}$$
where
    $$\Phi(s)=- \bu|_s \cdot \bn_s,$$
with extension by 0 to all of $\Omega$, and $\bu$ is the velocity solution on
    $$\Lambda(s) = \{(x,y,z) : b(x,y) < z < s(x,y)\}$$
\end{itemize}
\end{frame}

@@

\begin{frame}{what's needed for multigrid to work here?}

\begin{block}{viscous fluid layer geometry problem}
	$$\ip{\Phi(s)}{r-s} \ge \ip{a}{r-s} \quad \text{ for all } r \in \mathcal{K}$$
where
\begin{itemize}
\item $\mathcal{K} = \left\{r \in \mathcal{V} \,:\, r \ge b\right\}$
\item $\Phi(s)=- \bu|_s \cdot \bn_s$
\item $s$ is solution surface elevation
\item $\bu$ is Stokes solution on $\Lambda(s) = \{b < z < s\}$
\item $a$ (climate) and $b$ (bed elevation) are the input data
\end{itemize} 
\end{block}

\bigskip
\begin{itemize}
\item what is needed for scalable multilevel solutions?
    \begin{enumerate}
    \item<2>[1.] iterates must be admissible
    \item<2>[2.] global linearization must be avoided
    \item<2>[3.] smoother cost must be comparable to one residual    
    \end{enumerate}
\end{itemize}
\end{frame}

\section{multigrid approaches for VIs}

\begin{frame}[fragile]
\frametitle{Newton-multigrid for the classical obstacle problem}

\begin{itemize}
\item VIs are nonlinear problems, even for linear operators like $-\nabla^2$
\item Newton-multigrid is straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg \
    -pc_type mg -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] linear solver applies to inactive variables
        \begin{itemize}
        \item[] \texttt{rsls} = reduced space line search
        \end{itemize}
    \item[$\circ$] Newton step equations solved by CG with GMG V-cycles
    \end{itemize}
\item issue: the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] grid-dependent (growing) Newton iterations
    \end{itemize}
\end{itemize}

\medskip
\hspace{10mm} \includegraphics[height=0.2\textheight]{images/obstacle65.pdf}

\vspace{-20mm}
\hfill \includegraphics[height=0.25\textheight]{images/vi-newton-gmg-bad.png}
\end{frame}

\begin{frame}[fragile]
\frametitle{nested iteration}

@@

\vspace{2mm}
    \begin{itemize}
    \item[$\circ$] grid-independent Newton iterations
    \item[$\circ$] optimal $O(N^1)$ flops and time
    \item<2>[$\circ$] Chapter 12 example in my new book
    \end{itemize}

@@

\bigskip
\includegraphics[height=0.25\textheight]{images/vi-newton-gmg-good.png}

\vspace{-22mm}
\hspace{40mm} \includegraphics[height=0.45\textheight]{images/obstacle-flops-per-n.png} \qquad \includegraphics<2>[width=20mm]{images/frontcover.jpg}
\end{frame}

\begin{frame}{multigrid strategies for VIs}

\begin{itemize}
\item semi-smooth Newton can also yield mesh-independent iterations (Farrell et al.~2020)
\item other VI multilevel strategies:
    \begin{itemize}
    \item[$\circ$] projected FAS multigrid for linear CPs (Brandt \& Cryer, 1983)
    \item[$\circ$] monotone multigrid (Kornhuber, 1994)
    \item[$\circ$] multilevel constraint decomposition (Tai, 2003)
    \end{itemize}
\end{itemize}
\end{frame}

\section{multigrid approaches for nonlinear and \emph{nonlocal} VIs?}

\newcommand{\stacktwo}[2]{\begin{tabular}{c} #1 \\ #2 \end{tabular}}

\begin{frame}{multigrid strategies for nonlinear VIs: feature table}

{\small
\begin{tabular}{l|ccccc}
   & \stacktwo{admissible}{iterates} & \stacktwo{mesh-indep.}{rates}
       & \stacktwo{no global}{linearization}
         & \stacktwo{PETSc or}{Firedrake}  \\ \hline
RS NM                      & $\checkmark$ &  & & $\checkmark$ \\
\quad $+$ NI & $\checkmark$ & $\checkmark$ & & $\checkmark$ \\
SS NM             &              & $\checkmark$ & & $\checkmark$ \\
{\color{FireBrick} FASCD}  & $\checkmark$ & {\color{FireBrick} \textbf{?}} & $\checkmark$ & {\color{FireBrick} {\large $\checkmark$}}
\end{tabular}
}

\bigskip
{\scriptsize
\begin{center}
RS = reduced space, SS = semi-smooth, NM = Newton-multigrid, NI = nested iteration
\end{center}
}

\vspace{5mm}
\begin{itemize}
\item for the non-local fluid layer VI problem we need all 4 checked
\item we are trying-out a new algorithm,

{\color{FireBrick} FASCD = full approximation storage constraint decomposition}

    \begin{itemize}
    \item[$\circ$] Firedrake implementation \only<2>{\dots as of yesterday}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{constraint decomposition}

CD gloss

\begin{frame}{constraint decomposition}

dishonest attempt figure

\begin{itemize}
\item x
\end{itemize}

\includegraphics[width=0.5\textwidth]{images/icedecomp.png}
\end{frame}

\begin{frame}{full approximation storage constraint decomposition}

pseudocode
\end{frame}

\begin{frame}{evidence of mesh independence}

\begin{itemize}
\item 1D SIA
\end{itemize}

\includegraphics[width=0.5\textwidth]{images/sia-asymprates.pdf}
\end{frame}

\begin{frame}{2D SIA results (\emph{very fresh})}

\begin{itemize}
\item x
\end{itemize}

\includegraphics[width=0.5\textwidth]{images/fascd-firedrake-dome.png}
\end{frame}

@@

%{\scriptsize
{\notsotiny
\input{references.tex}
}
\end{frame}

@@

%{\scriptsize
{\notsotiny
\input{background.tex}
}
\end{frame}
